---
title: "prjsta426"
author: "hwartmann"
date: "November 30, 2016"
output:
  html_document:
    css: /home/hannes/Dropbox/ETH/sta426/prj/mystyle.css
  pdf_document: default
---
```{r "setup", include=FALSE}
knitr::opts_knit$set(root.dir = "/home/hannes/Dropbox/ETH/sta426/prj")  # with something else than `getwd()`
library(captioner)
library(citrus)
library(knitr)

load(file="/home/hannes/Dropbox/ETH/sta426/prj/img_23121000")


figs <- captioner(prefix="Figure")
tbls <- captioner(prefix="Table")

figs("eventcount_files","Total number of events per file")
capeventcount = figs("eventcount_files")
figs("box","The same data is plotted here as in Figure XX. We picked a random patient and looked at the boxplots for the mean marker intensity of 100 trials.", display=FALSE)
capbox = figs("box")
tbls("dataspread","Each files was downsampled with 5000 samples for 1000 times. Each time the mean marker intensity for the extracellular markers was calculated. Here we show the mean of 1000 means ± s.d.. Patient 1 has s.d. zero because the files contain less than 5000 events.")
capdataspread = tbls("dataspread")



```
##Introduction
Single cell measurements have a clear advantage over whole population measurements. In whole population measurements we measure the average value of a parameter over the total population in our sample. To answer questions that involve the study of specific subpopulation of cells this is not very helpful. With single cell measurements we are able to get values for each cell separately. Given single cell data from a sample we should be able to stratify the cells into its subpopulation and make some prediction towards a disease state, for example. 
The most common method to obtain single cell measurements is flow cytometry. For a long time fluorescent flow cytometry was very popular. With this method, in addition to physical properties such as shape, one can also measure the abundance of fluorescent markers within a cell. This technique limited by the spectral overlap, i.e. the overlap of the wavelength of the differently colored fluorescent used. This makes it very hard to distinguish a large number of parameters. In resent time flow cytometry has been coupled with mass spectrometry to allow the measurements of >40 inter and outer cellular parameters at the same time. Very briefly, in this method samples are incubated with antibodies carrying rare metals and are then vaporized in a time of flight mass spectrometer. The mass of the rare metals and their abundance is measured and replaced the fluorescent measurement of fluorescent flow cytometry.
Bruggner et al. have proposed a new automated method to detect stratifying signatures in cellular sub populations. They named their method citrus and it's available as an R package. Citrus combines unsupervised and supervised learning to select end-point i.e. classification significant features. First a predefined number of samples is randomly selected from each patient and the data is combined. This data is then clustered using hierarchical clustering. From these clusters citrus tried to extract relevant per-sample features. These features, together with the known end-point status i.e. healthy or diseased are used to train a regularized regression model.
Here in this project work we set out to explain the workings of citrus in more detail while following the same real world example that is used in the original paper. 

##Data
Bodenmiller et al. created a data set with a newly proposed mass-tag cellular barcoding (MCB) method. This method allows for very high dimensional data acquired with mass cytometry. More specifically they set out to measure signaling dynamics and cell-to-cell communication in peripheral blood mononuclear cells (PBMC). To this end 12 different experimental conditions were set up and 14 signaling nodes and 10 cell-surface markers were measured at different time points from 0-4h. Here we focus on one of those conditions mainly the cross-linking of the B-cell receptor (BCR)/Fc receptor (FcR). The data consist of 16 samples of eight healthy donors, one reference and one treated sample per patient.
```{r }

dataDirectory = "./data"
files = c("PBMC8_30min_patient1_BCR-XL.fcs","PBMC8_30min_patient1_Reference.fcs","PBMC8_30min_patient2_BCR-XL.fcs","PBMC8_30min_patient2_Reference.fcs","PBMC8_30min_patient3_BCR-XL.fcs","PBMC8_30min_patient3_Reference.fcs","PBMC8_30min_patient4_BCR-XL.fcs","PBMC8_30min_patient4_Reference.fcs","PBMC8_30min_patient5_BCR-XL.fcs","PBMC8_30min_patient5_Reference.fcs","PBMC8_30min_patient6_BCR-XL.fcs","PBMC8_30min_patient6_Reference.fcs","PBMC8_30min_patient7_BCR-XL.fcs","PBMC8_30min_patient7_Reference.fcs","PBMC8_30min_patient8_BCR-XL.fcs","PBMC8_30min_patient8_Reference.fcs")
fileList = data.frame(bcr=list.files(dataDirectory,pattern="BCR"),reference=list.files(dataDirectory,pattern="Reference"))

```
##Data sampling
For each patient there is a different number of events in the FCS files. As figure XX shows the number of events ranges from about 3000 in patient 1 to about 17'000 in patient 2. To prevent over representation of any one sample and to reduce computation time Citrus samples a user specified number of events per patient and then merges the data for clustering. The function sample() from the R base package is used to select 5000 random data points from the FCS files, if available. As we have tried to reproduce the exact data obtained bei Bruggner et al we found that this is not possible. We suspect this to be the result of downsampling the date. To get an idea how different the subsamples really are we decided to subsample 5000 data points for each data file 1000 times and calculate the mean and variance of the mean of the intensity of the markers.


```{r eval=FALSE}

eventcount = vector(mode="integer", length=16)
i = 1
for(f in files){
  fcsfile = citrus.readFCS(file.path(dataDirectory,f))
  print(dim(fcsfile))
  eventcount[i] = dim(fcsfile)[1]
  i = i +1
}
```
```{r fig.align='center',fig.cap=capeventcount}
par(mar=c(6,4,2,2)) #sets the bottom, left, top and right
barplot(eventcount,main="Event distribution", 
  names.arg=c("p1 BCR-XL","p1 ref","p2 BCR-XL","p2 ref","p3 BCR-XL","p3 ref","p4 BCR-XL","p4 ref","p5 BCR-XL","p5 ref","p6 BCR-XL","p6 ref","p7 BCR-XL","p7 ref","p8 BCR-XL","p8 ref"),las=2)

```
```{r eval=FALSE}

marker = c(3, 4, 9, 11, 12, 14, 21, 29, 31, 33)
maker_names = c("CD3","CD45","CD4","CD20","CD33","CD123","CD14","igM","HLA-DR","CD7")

data_analysis = matrix(data=NA, nrow=16, ncol=10)
colnames(data_analysis) = maker_names
rownames(data_analysis) = c("p1 BCR-XL","p1 Ref","p2 BCR-XL","p2 Ref","p3 BCR-XL","p3 Ref","p4 BCR-XL","p4 Ref","p5 BCR-XL","p5 Ref","p6 BCR-XL","p6 Ref","p7 BCR-XL","p7 Ref","p8 BCR-XL","p8 Ref")



avgintensity = matrix(data=NA, nrow=100, ncol=10)
colnames(avgintensity) = maker_names
avgintensity_plot = matrix(data=NA, nrow=100, ncol=10)
colnames(avgintensity_plot) = maker_names

k = 1
for(j in 1:8){
  #loop all bcr / ref, load files
    for(each in c("_BCR-XL.fcs","_Reference.fcs")){
      fcsfile = citrus.readFCS(file.path(dataDirectory,paste0("PBMC8_30min_patient",j,each)))
      #subsamples 1000 times
      for(i in 1:100){
        #if not patient 1, sample 5k, p1 has only ca. 3k samples
        if(j != 1){
          sample_fcsfile = fcsfile[sort(sample(1:nrow(fcsfile),5000)),] #1:numberofrowsin(fcsfile)
        }else{
          sample_fcsfile = fcsfile #1:numberofrowsin(fcsfile)
        }
        #loop each marker andstore mean
        for(h in 1:10){
          #avgintensity is 1000x10 i.e. 1000 subsamples with each 10 means for 10 markers
          avgintensity[i,h] = mean(sample_fcsfile@exprs[,marker[h]])
        }
      }
       if(j == 2 && each == "_BCR-XL.fcs"){
         avgintensity_plot = avgintensity
        
        
       }
      for(p in 1:10){
       data_analysis[k,p] = paste0(signif(mean(avgintensity[,p]), digits = 2),"±",signif(var(avgintensity[,p]), digits = 2))
      }
       k = k +1
      
    }
}

```
```{r,results='asis',fig.align='center',fig.cap=capbox}

 par(mfrow=c(5,2))
         par(mar=c(2,2,2,2))
         par(las = 1) # all axis labels horizontal
         for(x in 1:10){
            boxplot(as.data.frame(avgintensity_plot[,x]), main = paste("Patient 2 BCR-XL",marker_names[x]),
            horizontal = TRUE)
         }
```
```{r,results='asis'}

kable(data_analysis, caption = "Table1: Each files was downsampled with 5000 samples for 1000 times. Each time the mean marker intensity for the extracellular markers was calculated. Here we show the mean of 1000 means ± s.d.. Patient 1 has s.d. zero because the files contain less than 5000 events.",format = "html")
```

##Clustering of data by extracellular markers
One popular method for unsupervised learning is clustering. In clustering one can apply different algorithms to detect structure within a given data set and can build features based on cluster membership. Citrus uses Rclusterpp.hclust() also created by Bruggner et al. This function is an implementation of the Agglomerative Hierarchical Clustering algorithm. In hierarchical clustering smaller clusters are merged into larger clusters following a given distance function and linkage method. In Citrus the distance between markers is specified by the Euclidean distance and Wards linkage used as the agglomerative method. Wards methods minimizes the total within cluster variance.  
The Bodenmiller data is clustered by the intensity of the extracellular markers CD45,CD4,CD20,CD33,CD123,CD14,IgM,HLA-DR,CD7 and CD3. 
```{r eval=FALSE}
clusteringColumns = c("CD45(In115)Dd","CD4(Nd145)Dd","CD20(Sm147)Dd","CD33(Nd148)Dd","CD123(Eu151)Dd","CD14(Gd160)Dd","IgM(Yb171)Dd","HLA-DR(Yb174)Dd","CD7(Yb176)Dd","CD3(110:114)Dd")
citrus.clustering = citrus.cluster(citrus.combinedFCSSet,clusteringColumns,clusteringType = "hierarchical")
```
```{r eval=FALSE}
largeEnoughClusters1= citrus.selectClusters.minimumClusterSize(citrus.clustering,minimumClusterSizePercent=0.01)

hierarchyGraph = citrus.createHierarchyGraph(citrus.clustering,selectedClusters=largeEnoughClusters1)

clusterMedians = t(sapply(largeEnoughClusters1,citrus:::.getClusterMedians,clusterAssignments=citrus.clustering$clusterMembership,data=citrus.combinedFCSSet$data,clusterCols=clusteringColumns))

rownames(clusterMedians) = largeEnoughClusters1
colnames(clusterMedians) = clusteringColumns

citrus.plotClusteringHierarchy(outputFile="./output/clusterhierarchy.pdf",clusterColors=clusterMedians,graph=hierarchyGraph$graph,layout=hierarchyGraph$layout,plotSize=15, theme = 'white')
```
###Compare clustering with different subsamples
```{r eval=FALSE, echo=FALSE}
citrus.combinedFCSSet_1 = citrus.readFCSSet(dataDirectory,fileList, fileSampleSize = 5000)
citrus.clustering_1 = citrus.cluster(citrus.combinedFCSSet_1,clusteringColumns,clusteringType = "hierarchical")
citrus.combinedFCSSet_2 = citrus.readFCSSet(dataDirectory,fileList, fileSampleSize = 5000)
citrus.clustering_2 = citrus.cluster(citrus.combinedFCSSet_2,clusteringColumns,clusteringType = "hierarchical")
citrus.combinedFCSSet_3 = citrus.readFCSSet(dataDirectory,fileList, fileSampleSize = 5000)
citrus.clustering_3 = citrus.cluster(citrus.combinedFCSSet_3,clusteringColumns,clusteringType = "hierarchical")
citrus.combinedFCSSet_4 = citrus.readFCSSet(dataDirectory,fileList, fileSampleSize = 5000)
citrus.clustering_4 = citrus.cluster(citrus.combinedFCSSet_4,clusteringColumns,clusteringType = "hierarchical")
citrus.combinedFCSSet_5 = citrus.readFCSSet(dataDirectory,fileList, fileSampleSize = 5000)
citrus.clustering_5 = citrus.cluster(citrus.combinedFCSSet_5,clusteringColumns,clusteringType = "hierarchical")

#make a list of the clusterings to loop though them
subsamples = list(citrus.combinedFCSSet_1,citrus.combinedFCSSet_2,citrus.combinedFCSSet_3,citrus.combinedFCSSet_4,citrus.combinedFCSSet_5)
clusterings = list(citrus.clustering_1,citrus.clustering_2,citrus.clustering_3,citrus.clustering_4,citrus.clustering_5)

total_count_1 = vector(mode ="integer", length=16)
total_count_2 = vector(mode ="integer", length=16)
total_count_3 = vector(mode ="integer", length=16)
total_count_4 = vector(mode ="integer", length=16)
total_count_5 = vector(mode ="integer", length=16)

total_patient_count = list(total_count_1,total_count_2,total_count_3,total_count_4,total_count_5)

outermostclusters_1 = c()
outermostclusters_2 = c()
outermostclusters_3 = c()
outermostclusters_4 = c()
outermostclusters_5 = c()

outermostcluster_list = list(outermostclusters_1,outermostclusters_2,outermostclusters_3,outermostclusters_4,outermostclusters_5)

#for each clustering 1:5
i = 1
for(each_c in clusterings){
  #find the clusters containing at leat 5% of the total data
  largeenoughclusters_a =citrus.selectClusters.minimumClusterSize(each_c,minimumClusterSizePercent=0.05)
  #find bottom clusters
  for(each_subc in largeenoughclusters_a){
    decendants = citrus.getClusterDecendants(each_subc,each_c$clustering$merge)
    decendantsinLEC = decendants %in% largeenoughclusters_a
    if(length()){
      outermostcluster_list[[i]] = c(outermostcluster_list[[i]],each_subc)
    }
  }
  i = i +1
}
  
for(j in 1:5){
    #for each of the bottom clusters per clustering
  for(each_subc in outermostcluster_list[[j]]){
      #find the events contained in the bottom clusters
      eventsincluster = each_c$clusterMembership[[each_subc]]
  }
   #for each event per cluster, look up its patient id and increment the counter
   for(each_event in eventsincluster){
     total_patient_count[[j]][subsamples[[1]]$data[each_event,37]] = total_patient_count[[j]][subsamples[[1]]$data[each_event,37]] +1
   }
}
    




#some plots


#reduced_counts = Reduce("+",total_patient_count)
treated_count = sum(reduced_counts[1:8])
healthy_count = sum(reduced_counts[9:16])


#healthy vs. sick
barplot(,main="Event distribution", 
  names.arg=c("p1 BCR-XL","p1 ref","p2 BCR-XL","p2 ref","p3 BCR-XL","p3 ref","p4 BCR-XL","p4 ref","p5 BCR-XL","p5 ref","p6 BCR-XL","p6 ref","p7 BCR-XL","p7 ref","p8 BCR-XL","p8 ref"),las=2)

```

###Hierarchygraph

```{r}

```


##features
```{r eval=FALSE, echo=FALSE}
# Vector of parameters to calculate medians for
functionalColumns = c("pNFkB(Nd142)Dd", "pp38(Nd144)Dd", "pStat5(Nd150)Dd", "pAkt(Sm152)Dd", "pStat1(Eu153)Dd", "pSHP2(Sm154)Dd", "pZap70(Gd156)Dd", "pStat3(Gd158)Dd", "pSlp76(Dy164)Dd", "pBtk(Er166)Dd", "pPlcg2(Er167)Dd", "pErk(Er168)Dd","pLat(Er170)Dd", "pS6(Yb172)Dd")

# Build features
abundanceFeatures = citrus.calculateFeatures(citrus.combinedFCSSet, clusterAssignments=citrus.clustering$clusterMembership,clusterIds=largeEnoughClusters5)

medianDifferenceFeatures = citrus.calculateFeatures(citrus.combinedFCSSet,
                                                clusterAssignments=citrus.clustering$clusterMembership,
                                                clusterIds=largeEnoughClusters5,
                                                featureType="medians",
                                                medianColumns=functionalColumns,
                                                conditions=c("reference","bcr"))
medianFeatures = citrus.calculateFeatures(citrus.combinedFCSSet,
                                                clusterAssignments=citrus.clustering$clusterMembership,
                                                clusterIds=largeEnoughClusters5,
                                                featureType="medians",
                                                medianColumns=functionalColumns
                                                )

```
##Plotting hierarchycal Graphs + differential features
```{r eval=FALSE, echo=FALSE}
largeEnoughClusters1= citrus.selectClusters.minimumClusterSize(citrus.clustering,minimumClusterSizePercent=0.01)

hierarchyGraph = citrus.createHierarchyGraph(citrus.clustering,selectedClusters=largeEnoughClusters1)

clusterMedians = t(sapply(largeEnoughClusters1,
                          citrus:::.getClusterMedians,
                          clusterAssignments=citrus.clustering$clusterMembership,
                          data=citrus.combinedFCSSet$data,
                          clusterCols=clusteringColumns))

rownames(clusterMedians) = largeEnoughClusters1
colnames(clusterMedians) = clusteringColumns

citrus.plotClusteringHierarchy(outputFile="./output/clusterhierarchy.pdf",
                               clusterColors=clusterMedians, 
                               graph=hierarchyGraph$graph,
                               layout=hierarchyGraph$layout,
                               plotSize=hierarchyGraph$plotSize)

##plot featurs highlated in hierarchygraph

#get ps6 values coupled withh fileID
ps6dataperevent = citrus.combinedFCSSet$data[,c(30,37)]
#replace fileID with endpoint label ref = 0, bcr = 1
bcr = c(1,2,3,4,5,6,7,8)
ref = c(9,10,11,12,13,14,15,16,17,18)

for(i in 1:dim(ps6dataperevent)[1]){
  if(ps6dataperevent[i,2] %in% bcr){
    ps6dataperevent[i,2] = 1
  }else{
    ps6dataperevent[i,2] = 0
  }
}

datafordiff = vector(mode = "list", length = length(largeEnoughClusters1))

#loop though each cluster and select all events in that cluster
j = 1

#citrus.cluster$clustermembership is cluster X events NOT events X cluster

for(eachcluster in largeEnoughClusters1){
  print(citrus.clustering$clusterMembership[[eachcluster]])
  j = j +1
}



# for(eachcluster in largeEnoughClusters1){
#   print(j) 
#   eventlist = 0
#   for(i in 1:length(citrus.clustering$clusterMembership)){
#     #if we find the cluster listed for the event, we asign the index of the event to the cluster
#    
#     if(eachcluster %in% citrus.clustering$clusterMembership[i][[1]]){
# 
#       eventlist = c(eventlist,i)
#       print(eventlist)
#     }
#     
#   }
#   datafordiff[j] = eventlist
#   j = j +1
# }


y = matrix(data=c(largeEnoughClusters1,), nrow=largeEnoughClusters1, ncol=8)
 


grp <- rep(0:1,each=nSamples/2)





#later get the cluster where the most sig feature is 
# Features to highlight
featureClusterMatrix = data.frame(cluster=c(19992,19978,19981,19987,19983,19973),feature=rep(c("Property 1","Property 2"),each=3))

# Plot features in clustering hierarchy
citrus.plotHierarchicalClusterFeatureGroups(outputFile="./output/testfeat.pdf",
                                              featureClusterMatrix,
                                              graph=hierarchyGraph$graph,
                                              layout=hierarchyGraph$layout,
                                              plotSize=hierarchyGraph$plotSize)
```

#calculate regularization thresholds
```{r}
regularizationThresholds = citrus.generateRegularizationThresholds(medianFeatures,
                                                                    labels,
                                                                    modelType="pamr", 
                                                                    family="classification")
```

#calculate cv error rates
```{r}
thresholdCVRates = citrus.thresholdCVs.quick("pamr",
                                             medianFeatures,
                                             labels,
                                             regularizationThresholds,
                                             family="classification")
```

#calculate cvminima
```{r}
cvMinima = citrus.getCVMinima("pamr",thresholdCVRates, fdrRate = 0.01)
```

#Build final model
```{r}
finalModel = citrus.buildEndpointModel(medianFeatures,
                                        labels, 
                                        family="classification",
                                        type="pamr",
                                        regularizationThresholds)

#endpointModel = citrus.buildEndpointModel(abundanceFeatures,labels)
```

#Extracting Model Features.
```{r}
modelfeatures = citrus.extractModelFeatures(cvMinima,finalModel,
                                            medianFeatures)
```{r}
```
```{r}
```
```{r}
```
```{r}
```

